Density-dependent selection and the limits of relative fitness

### 2017-12-15 manuscript.tex
l

Evolutionary biologists frequently describe selection using relative fitness, and relative fitness models are foundational in population genetics. Yet the population ecological basis of relative fitness is poorly understood, and the classical ecology literature on selection in crowded populations seems to be incompatible with widely-used relative fitness models such as the Wright-Fisher. Here we develop a generalization of the Wright-Fisher model in which density depends dynamically on the demographic rates of the types present. We then explore the population ecology of relative fitness using this model as a reference point. Although density-dependent selection can cause relative fitness to break down, relative fitness models are fairly robust in populations close to demographic equilbrium. In particular, relative fitness only breaks down if selection is strong and density-dependent, and density is also selection-dependent. Our generalized Wright-Fisher model clearly distinguishes between three demographic parameters, each of whose behavior is, on its own, correctly described by relative fitness models. In contrast, the classical literature on density-dependent selection confounds them. We argue that selection-independent density is ecologically plausible for many species given the prevalence of reproductive excesses, an important aspect of selection omitted in many ecological models. Our model also offers a natural alternative to relative fitness when the latter is untenable, as is likely the case far from demographic equilibrium.

@@

The difficulty with Eq.~\eqref{eq:canonical} arises in crowded populations. Since crowded and uncrowded conditions are so different, $s$ will often depend on density. Eq.~\eqref{eq:canonical} is then no longer a complete description of selection --- we would also need to specify a model for density. Note that frequency-dependent selection does not raise similar problems; Eq.~\eqref{eq:canonical} is still a complete description of selection even if its behavior is more complicated compared to the constant-$s$ case. Standard population genetics evades the issue of density-dependent selection by simply assuming that total population density $N$ has reached its equilibrium value, which is assumed to be a fixed constant. The selection coefficient $s$ now parameterizes the rate at which selection changes relative frequencies, but no longer corresponds to differences in intrinsic growth rates $r$.

Counter to the assumption of constant $N$, MacArthur famously showed that when population growth is density-regulated, selection in crowded populations is intimately connected to the ability to keep growing at higher densities than other types can tolerate \citep{macarthur_1967}. The classic example is the logistic model, where the type with the greatest saturation population density ``$K$'' excludes the others (Fig.~\ref{fig:Ksel}a). Similarly, the ``$R^*$ rule'', a central tenet of resource competition theory, states that when growth is limited by a single homogeneous consumable resource, the type able to deplete the resource to the lowest equilibrium density $R^*$ excludes the others \citep{grover_1997}. Differences in $R^*$ will often entail differences in saturation density. The Lotka-Volterra competition model also couples selection in crowded populations to changes in $N$ except in a few special cases (we return to this in section ``$K$-selection versus relative fitness''). In these examples, both $N$ and $s$ change during, and as a result of, adaptive sweeps. It would therefore seem that the ubiquitous constant-$N$, relative fitness description of selection is incompatible with a huge class of population ecological processes driving selection (Fig.~\ref{fig:Ksel}b).

@@

The relative fitness description has been justified in broadly two different ways for crowded populations (we do not discuss Wagner's [\citeyear{wagner_2010}] measure-theoretical justification, which is explicitly independent of population biology and thus falls outside of our scope). The first is to assume that selection is independent of density but still allow density to be affected by selection \citep[pp. 468]{barton_2007} \citep{prout_1980}. While this  allows us to relax the assumption of constant $N$, it does not address the problem that $s$  can, in reality, depend on density. In the examples from the previous paragraph, selection is density-dependent; indeed, the type-specific responses to density are at the center of MacArthur's argument and the density-dependent selection literature that grew out of it \citep{roughgarden_1979}.

Second, constant $N$ and $s$ can both be seen as a short-term linear approximation \cite[pp. 277]{ewens_2004}. That is, within a sufficiently short time frame, $N$ and $s$ can be treated as constant. Provided that selection is sufficiently weak and stable over time, this ``short-term''  assumption is not a major restriction. Yet it is increasingly recognized that selection is not always weak, that it can fluctuate considerably over time, and that $N$ can vary by orders of magnitude over a few generations as a routine feature of a population's ecology \citep{messer_2016}. These are not rare exceptions, but occur widely in nature and the lab, including in wild \textit{Drosophila} \citep{bergland_14}. Nevertheless, relative fitness models are the foundation for much of the population genetic literature, and are still widely used without considering the ``short-term'' restriction or the lack of integration with population ecology \citep{mallet_2012}. Thus, it is important to understand the population ecological basis of relative fitness models, both to gain insight into their domain of applicability, and as part of the broader challenge of synthesizing ecology and evolution.

Another issue with the constant-$N$ relative fitness description of selection is that it precludes consideration of longer-term aspects of the interplay between evolution and ecology such as population extinction. [will make this sound less dismissive] Adaptive dynamics currently provides a powerful framework for addressing the complex feedbacks between evolutionary change and population density, including but not limited to extinction \citep{diekmann_2004}. However, the focus of adaptive dynamics is trait evolution rather than the underlying genetics, and in particular, selective sweeps are typically subsumed into effectively-instantaneous ``trait substitutions''. We emphasize that our focus here is the description of the time-dependent process by which selection changes allele frequencies, which is particularly critical for making sense of evolution at the genetic level for which we now have abundant data.

Here we analyze the population ecology of relative fitness using a novel model of density-dependent population growth based on territorial contests. Rather attempting to make sense of relative fitness in existing standard models of population growth mentioned above (e.g. \citep{kimura1969natural,mallet_2012}), we instead do the reverse, and attempt to make population ecological sense of the widely-used Wright-Fisher, constant-$N$, relative fitness model. Our starting point is the classic lottery model of territorial contest \citep{sale_77,chesson_1981}. Like the Wright-Fisher model, the classic lottery assumes a saturated population with constant $N$, and fitness involves a product of fertility and juvenile viability \citep[pp. 185]{crow_1970}. Unlike the Wright-Fisher model, generations can overlap in the lottery model. We generalize the lottery model to allow population density to vary, creating a density-dependent generalization of the Wright-Fisher model with overlapping generations.

We show that when this model reaches a demographic steady-state, the constant-$N$, relative fitness picture emerges. Futhermore, we show that our model is entirely consistent with MacArthur's analysis of selection in crowded populations. In particular, we emphasize that MacArthur's argument does not justify the widespread intuition that selection in crowded environments is necessarily connected to achieving greater densities. This intuition is largely an artifact of the models historically used in the density-dependent selection literature, which ignore the existence of relative contests.

Our first task is to analytically extend the classic lottery model to correctly account for low density behavior (sections ). We then...[BLAH]

@@

\subsection*{Model assumptions and definitions}

We assume that reproductively mature individuals (``adults'') each require their own territory to survive and reproduce (Fig.~\ref{fig:lottery}). All territories are identical, and the total number of territories is $T$. Time advances in discrete iterations, each representing the time from birth to reproductive maturity. In a given iteration the number of adults of the $i$'th type will be denoted $n_i$, the total number of adults by $N=\sum_i n_i$, and the number of unoccupied territories by $U=T-N$.

We assume that the $n_i$ are large enough that stochastic fluctuations in the $n_i$ (``drift'') can be ignored. In particular, we do not evaluate the initial stochastic behaviour of mutant lineages while they are at low abundance. We derive equations for the expected change in the $n_i$ over time, leaving the evaluation of drift for future work.

%When considering new mutations, we therefore restrict our attention to the earliest (lowest $n_i$) deterministic behavior of mutant lineages (the transition to deterministic growth occurs at an abundance $n_i$ of order equal to their inverse expected absolute growth rate; \citealt{uecker_2011}).

@@

Each iteration, adults produce new offspring (``propagules''), $m_i$ of which disperse to unoccupied territories. We assume that adults cannot be ousted from their territories, so that $m_i$ only includes propagules landing on unoccupied territories. Propagules disperse at random over the unoccupied territories, regardless of distance from their parents, and independently of each other. There is no interaction between propagules (e.g. avoidance of territories crowded with propagules). Loss of propagules during dispersal is subsumed into $m_i$. We assume that each adult produces a constant number $b_i$ of successfully dispersing propagules; the loss of propagules due to dispersal to occupied territories then implies $m_i=b_i(1-N/T)n_i$. Note that due to our assumption of uniform dispersal, the parameter $b_i$ can be thought of as a measure of ``colonization ability'', which combines fecundity and dispersal ability \citep{levins_71,tilman_94,bolker_99}.

The number of individuals of the $i$'th type landing in any particular territory is denoted $x_i$. We assume that $x_i$ follows a Poisson distribution $p_i(x_i)=l_i^{x_i} e^{-l_i}/x_i!$, where $l_i=m_i/U$ is the mean territorial propagule density. This approximation becomes exact when the $n_i$ are large enough that drift in $n_i$ can be ignored (Appendix A).

When multiple propagules land on the same territory, the victor is determined by lottery competition: type $i$ wins a territory with probability $c_i x_i/\sum_j c_j x_j$, where $c_i$ is a constant representing relative competitive ability (Fig. \ref{fig:lottery}). We expect that a fraction $p_1(x_1)\ldots p_G(x_G)$ of the $U$ unoccupied territories will have the propagule composition $x_1,\ldots,x_G$. Type $i$ is expected to win $c_i x_i/\sum_j c_j x_j$ of these. Ignoring fluctuations about these two expectations (due to our no-drift, large $T$, large $n_i$ approximation), type $i$'s territorial acquisition is given by
\begin{equation}
\Delta_+ n_i=U\sum_{x_1,\ldots,x_G} \frac{c_i x_i}{\sum_j c_j x_j} p_1(x_1)\ldots p_G(x_G) \label{eq:growthsumuncoupled}
\end{equation}
in our extended lottery model, where the sum only includes territories with at least one propagule present.

Finally, we assume that mortality only occurs in adults (setting aside the juvenile deaths implicit in territorial contest), and at a constant, type-specific per-capita rate $0\leq d_i\leq 1$, so that the overall change in type abundances is
\begin{equation}
\Delta n_i=\Delta_+ n_i-d_i n_i. \label{eq:delttot}
\end{equation}

@@

In the classic lottery model \citep{chesson_1981}, unoccupied territories are assumed to be saturated with propagules from every type $l_i\gg 1$. From the law of large numbers, the composition of propagules in each territory will then not deviate appreciably from the mean composition $l_1,l_2,\ldots,l_G$ ($G$ is the number of types present), and so the probability that type $i$ wins any particular unoccupied territory is approximately $c_i l_i/\sum_j c_j l_j$. Then the numbers of territories won by each type $\Delta_+ n_1,\Delta_+ n_2,\ldots,\Delta_+ n_G$ follow a multinomial distribution with $U$ trials and success probabilities $\frac{c_1 l_1}{\sum_j c_j l_j},\frac{c_2 l_2}{\sum_j c_j l_j},\ldots,\frac{c_G l_G}{\sum_j c_j l_j}$, respectively. Type $i$ is expected to win $c_i l_i/\sum_j c_j l_j$ of the $U$ available territories, and deviations from this expected outcome are small (since $T$ is large by assumption), giving 
\begin{equation}
\Delta_+ n_i=\frac{c_i l_i}{\sum_j c_j l_j}U=\frac{c_i l_i}{\overline{c}L}U, \label{eq:lottery}
\end{equation}
where $\overline{c}=\sum_j c_j m_j/M$ is the mean propagule competitive ability for a randomly selected propagule, $L=M/U$ is the total propagule density and $M=\sum_j m_j$ is the total number of propagules. In section ``Analytical approximation of the density-dependent lottery'', we derive a generalization of Eq.~\eqref{eq:lottery} that accommodates arbitrary propagule densities $l_i$.

There is a close connection between the classic lottery model and the Wright-Fisher model of genetic drift \citep{svardal_2015}. In the Wright-Fisher model, type abundances are sampled each generation from a multinomial distribution with success probabilities $w_i n_i/\sum_j w_j n_j$, where $w$ is relative fitness and the $n_i$ are  type abundances in the preceding generation. Population size $N$ remains constant. This is equivalent to the classic lottery model with non-overlapping generations ($d_i=1$ for all $i$) and relative fitness given by $w_i=b_i c_i$ i.e. a product of fecundity and viability \citep[pp. 185]{crow_1970}. Thus, the classic lottery model is essentially the Wright-Fisher model extended to allow overlapping generations, but ignoring drift. This means that our extension of the classic lottery model to arbitrary densities represents a density-dependent generalization of the Wright-Fisher model (we also do not consider drift here).

@@

Eq.~\eqref{eq:master} takes a much simpler form if all types are competitively equivalent ($c_i=c$),
\begin{equation}
\Delta_+ n_i = \frac{l_i}{L}(1-e^{-L})U. \label{eq:masterequalc}
\end{equation}
Here $1-e^{-L}$ is the fraction of territories that receive at least one propagule under Poisson dispersal, $(1-e^{-L})U$ is the total number of territories gained, and type $i$ receives $l_i/L$ of these. Total population density thus grows according to
\begin{equation}
\Delta N=(1-e^{-L})U-\sum_i d_i n_i \label{eq:Nmaster}
\end{equation}

\subsection*{$K$-selection versus relative fitness}

We now compare the density-dependent lottery model from the previous section with MacArthur's analysis of selection in crowded environments \citep{macarthur_1967}. MacArthur considers a population with two types that have densities $n_1$ and $n_2$ subject to density-dependent growth described by
\begin{equation}
\frac{d n_1}{d t}=f_1(n_1,n_2)\qquad\frac{d n_2}{d t}=f_2(n_1,n_2). \label{eq:macgeneral}
\end{equation}
The environment is assumed to remain constant apart from the type densities. The functions $f_1$ and $f_2$ must decline to zero if $n_1$ or $n_2$ are sufficiently large, because no population has unlimited resources. This defines the nullclines $f_1(n_1,n_2)=0$ and $f_2(n_1,n_2)=0$ in $(n_1,n_2)$ space. The outcome of selection is then determined by the relationship between these nullclines. Specifically, a type will be excluded if its nullcline is completely contained in the region bounded by the other type's nullcline. In other words, for a type to have the possibility of persisting, it must be able to keep growing to higher densities than the other type can tolerate in some region of $(n_1,n_2)$ space (Fig.~\ref{fig:Ksel}a).

@@

Thus, when MacArthur concludes that  ``fitness is $K$'' in crowded populations \citep[pp. 149]{macarthur_1967}, the meaning is that selection either favors the ability to keep growing at ever higher densities (moving a type's own nullcline outwards), or the ability to suppress the growth of competitors at lower densities (moving the nullcline of competitors inwards) \citep{gill_1974}. This general idea applies even if the nullclines are nonlinear to such an extent that the ``$K$'' values themselves do not give much information about the regions of high-density growth.

@@

It is obvious from Eq.~\eqref{eq:LV} that selection can favor a superior competitor in a crowded population even if its saturation density is the same as, or lower than that of the other types present. However, note that the Lotka-Volterra model still closely couples selection to population density \citep{smouse_1976}. Fig.~\ref{fig:LVvslottery}a shows Lotka-Volterra selection between two types with the same saturation density ($\alpha_{11}=\alpha_{22}$, $\alpha_{21}>\alpha_{12}$). Even though the initial and final densities of a sweep are the same, density is not constant over a sweep. Only a highly restricted subset of $r$ and $\alpha$ values will keep $N$ constant over a selective sweep (further details in Appendix C). Intuitively, for one type to exclude another with the same saturation density, competitive suppression of growth between types must be stronger than competitive suppression of growth within types, causing a dip in $N$ over the sweep.

@@

@@

We are now in a position to analyze the validity of Eq.~\eqref{eq:canonical} more explicitly. In the previous section we showed that selection and the regulation of population density can be completely independent of each other even if population growth is density-regulated, and moreover that MacArthur's general argument \citep{macarthur_1967} never precluded this possibility. Nevertheless, selection and density regulation \textit{are} intimately connected in widely used models of population growth. To understand why this poses a problem for Eq.~\eqref{eq:canonical}, consider the simple birth-death model \cite[pp. 20]{kostitzin_1939} 
\begin{equation}
\frac{d n_i}{dt}=(b_i -\delta_iN) n_i \label{eq:simplebirthdeath}
\end{equation}
where $\delta_i$ is the per-capita increase in type $i$'s mortality rate due to crowding (for simplicity, there are no deaths when uncrowded). Then, starting from a monomorphic population, the frequency of an adaptive $\delta$-variant $\delta_i\rightarrow \delta_i(1-\epsilon)$ obeys 
\begin{equation}
\frac{d p_i}{dt}=\epsilon \delta_i N p_i(1-p_i). \label{eq:Ndependentsweep}
\end{equation}
The selection coefficient $s=\epsilon \delta_i N$ thus depends on density (compare with Model III in \cite{kimura1969natural}). On the other hand, the frequency of an adaptive $b$-variant $b_i\rightarrow b_i(1+\epsilon)$ will exactly obey Eq.~\eqref{eq:canonical} with $s=\epsilon b_i$, independent of density.

@@

@@

>>>>

Evolutionary biologists frequently describe selection using relative fitness, and relative fitness models are foundational in population genetics. Yet the population ecological basis of relative fitness is poorly understood, and the classical ecology literature on selection in crowded populations seems to be incompatible with widely-used relative fitness models such as the Wright-Fisher. Here we develop a generalization of the Wright-Fisher model in which density depends dynamically on the demographic rates of the types present. We then explore the population ecology of relative fitness using this model as a reference point. Although density-dependent selection can cause relative fitness to break down, relative fitness models are fairly robust in populations close to demographic equilibrium. In particular, relative fitness only breaks down if selection is strong and density-dependent, and density is also selection-dependent. Our generalized Wright-Fisher model clearly distinguishes between three demographic parameters, each of whose behavior is, on its own, correctly described by relative fitness models. In contrast, the classical literature on density-dependent selection confounds them. We argue that selection-independent density is ecologically plausible for many species given the prevalence of reproductive excesses, an important aspect of selection omitted in many ecological models. Our model also offers a natural alternative to relative fitness when the latter is untenable, as is likely the case far from demographic equilibrium.

@@

The difficulty with Eq.~\eqref{eq:canonical} arises in crowded populations. Since crowded and uncrowded conditions are so different, it seems likely that $s$ will often depend on density (although empirical verification has proved challenging \citep{travis_2013}). Eq.~\eqref{eq:canonical} is then no longer a complete description of selection --- we would also need to specify a model for density. Note that frequency-dependent selection does not raise similar problems; Eq.~\eqref{eq:canonical} is still a complete description of selection even if its behavior is more complicated due to $s$ depending on frequency. Standard population genetics evades the issue of density-dependent selection by simply assuming that total population density $N$ has reached its equilibrium value, which is assumed to be a fixed constant. The selection coefficient $s$ now parameterizes the rate at which selection changes relative frequencies, but no longer corresponds to differences in intrinsic growth rates $r$.

Counter to the assumption of constant $N$, MacArthur famously argued that when population growth is density-regulated, selection in crowded populations is intimately connected to the ability to keep growing at higher densities than other types can tolerate \citep{macarthur_1967}. The classic example is the logistic model, where the type with the greatest saturation population density ``$K$'' excludes the others (Fig.~\ref{fig:Ksel}a). Similarly, the ``$R^*$ rule'', a central tenet of resource competition theory, states that when growth is limited by a single homogeneous consumable resource, the type able to deplete the resource to the lowest equilibrium density $R^*$ excludes the others \citep{grover_1997}. Differences in $R^*$ will often entail differences in saturation density. The Lotka-Volterra competition model also couples selection in crowded populations to density except in special cases \citep{smouse_1976,mallet_2012}. In these examples, both $N$ and $s$ change during, and as a result of, adaptive sweeps. It would therefore seem that the ubiquitous constant-$N$, relative fitness description of selection is incompatible with a huge class of population ecological processes driving selection (Fig.~\ref{fig:Ksel}b).

@@

In light of these difficulties, the relative fitness description has been justified in broadly three different ways for crowded populations (we do not discuss Wagner's [\citeyear{wagner_2010}] measure-theoretical justification, which is independent of population biology). The first is to simply assume that selection is density-independent but relax the assumption of constant $N$ by allowing density to change as a result of selective sweeps \citep[pp. 468]{barton_2007} \citep{prout_1980}. Obviously this  does not address the problem that $s$ can, in reality, depend on density. Type-specific responses to density are at the center of MacArthur's argument and the density-dependent selection literature that grew out of it \citep{roughgarden_1979}.

The second justification, which grew out of a controversy over Haldane's ``cost of selection'', is to appeal to the existence of a ``reproductive excess'' of juveniles that are more fragile than their adult counterparts \citep{turner1968population,kimura1969natural,nei1971fertility}. Selection can then be concentrated at the juvenile phase, uncoupling selection from population density at the adult phase unless it is so strong that the reproductive excess is depleted. This justifies the use of Eq.~\eqref{eq:canonical} because, for a population in demographic equilibrium, selective sweeps do not affect density, and so it does not matter if selection is density-dependent. However, this reproductive excesses literature was focused on the evaluation of genetic load, and was poorly integrated with population ecology. \cite{kimura1969natural} took constant $N$ as a requirement and then introduced some mathematical variants of the logistic model that satisfy this requirement. \cite{nei1971fertility} proposed a model with an explicit representation of reproductive excess, but used an unusual model of competition based on pair-wise interactions which was only defined for at most two different types. As a result, the role of reproductive excesses in justifying Eq.~\eqref{eq:canonical} is still largely verbal.

Thirdly, Eq.~\eqref{eq:canonical} can justified as a short-term linear approximation \cite[pp. 277]{ewens_2004}. That is, within a sufficiently short time frame, $N$ and $s$ can be treated as constant. Provided that selection is sufficiently weak and stable over time, this ``short-term''  assumption is not a major restriction. Yet it is increasingly recognized that selection is not always weak, that it can fluctuate strongly over time, and that $N$ can vary by orders of magnitude over a few generations as a routine feature of a population's ecology \citep{messer_2016}. These are not rare exceptions, but occur widely in nature and the lab, including in wild \textit{Drosophila} \citep{bergland_14}.

Nevertheless, relative fitness models are the foundation for much of the population genetic literature, and are still widely used without considering the ``short-term'' restriction or the lack of integration with population ecology \citep{mallet_2012}. Thus, it is important to understand the population ecological basis of relative fitness models, both to gain insight into their domain of applicability, and as part of the broader challenge of synthesizing ecology and evolution.

Here we analyze the population ecology of relative fitness using a novel model of density-dependent population growth based on territorial contests. Rather attempting to make sense of relative fitness in existing standard models of population growth mentioned above (e.g. \citep{kimura1969natural,mallet_2012}), we instead do the reverse, and attempt to make population ecological sense of the widely-used Wright-Fisher, constant-$N$, relative fitness model. Our starting point is the classic lottery model of territorial contest \citep{sale_77,chesson_1981}. Like the Wright-Fisher model, the classic lottery assumes a saturated population with constant $N$, and fitness involves a product of fertility and juvenile viability \citep[pp. 185]{crow_1970}. Unlike the Wright-Fisher model, generations can overlap in the lottery model. In the ``Model'' section and ``Analytical approximation of the density-dependent lottery'' part of the results section, we generalize the lottery model to allow population density to vary, creating a variable-density version of the Wright-Fisher model with overlapping generations.

In the ``$K$-selection and selection-dependent density'' section, we show that when our model reaches a demographic steady-state, the constant-$N$, relative fitness picture emerges.

Furthermore, we show that our model is entirely consistent with MacArthur's analysis of selection in crowded populations.

@@

\subsection*{Assumptions and definitions}

We assume that reproductively mature individuals (``adults'') each require their own territory to survive and reproduce (Fig.~\ref{fig:lottery}). All territories are identical, and the total number of territories is $T$. Time advances in discrete iterations, each representing the time from birth to reproductive maturity. In a given iteration, the number of adults of the $i$'th type will be denoted by $n_i$, the total number of adults by $N=\sum_i n_i$, and the number of unoccupied territories by $U=T-N$. We assume that the $n_i$ are large enough that stochastic fluctuations in the $n_i$ (``drift'') can be ignored (with $T$ also assumed large to allow for small type densities $n_i/T$).

@@

Each iteration, adults produce new offspring (``propagules''). These disperse at random, regardless of distance from their parents, and independently of each other (e.g. there is no avoidance of territories crowded with propagules). We assume that adults cannot be ousted by juveniles, so that propagules landing on occupied territories are doomed. We assume that each adult from type $i$ produces a constant number $b_i$ of successfully dispersing propagules; the number of propagules dispersing to unoccupied territories is then given by $m_i=b_in_iU/T$. Note that due to our assumption of uniform dispersal, the parameter $b_i$ can be thought of as a measure of ``colonization ability'', which combines fecundity and dispersal ability \citep{levins_71,tilman_94,bolker_99}.

The number of propagules of the $i$'th type landing in any particular territory is denoted $x_i$. We assume that $x_i$ follows a Poisson distribution $p_i(x_i)=l_i^{x_i} e^{-l_i}/x_i!$, where $l_i=m_i/U$ is the mean territorial propagule density. This is strictly only an approximation of random dispersal, but it is an excellent approximation given our assumption that the $n_i$ are large enough that drift can be ignored (Appendix A).

When multiple propagules land on the same territory, the victor is determined by lottery competition: type $i$ wins a territory with probability $c_i x_i/\sum_j c_j x_j$, where $c_i$ is a constant representing relative competitive ability (Fig. \ref{fig:lottery}). We expect that a fraction $p_1(x_1)\ldots p_G(x_G)$ of the $U$ unoccupied territories will have the propagule composition $x_1,\ldots,x_G$. Type $i$ is expected to win $c_i x_i/\sum_j c_j x_j$ of these. Ignoring fluctuations about these two expectations (due to our no-drift, large $n_i$, large $T$ approximation), type $i$'s territorial acquisition is given by
\begin{equation}
\Delta_+ n_i=U\sum_{x_1,\ldots,x_G} \frac{c_i x_i}{\sum_j c_j x_j} p_1(x_1)\ldots p_G(x_G), \label{eq:growthsumuncoupled}
\end{equation}
where the sum only includes territories with at least one propagule present.

Finally, we assume that adult mortality only occurs in adults present at the start of the  iteration, and at a constant, type-specific per-capita rate $0\leq d_i\leq 1$. Thus, the overall change in type abundances is
\begin{equation}
\Delta n_i=\Delta_+ n_i-d_i n_i. \label{eq:delttot}
\end{equation}

@@

In the classic lottery model \citep{chesson_1981}, unoccupied territories are assumed to be saturated with propagules from every type $l_i\gg 1$. From the law of large numbers, the composition of propagules in each territory will then not deviate appreciably from the mean composition $l_1,l_2,\ldots,l_G$ ($G$ is the number of types present), and so the probability that type $i$ wins any particular unoccupied territory is approximately $c_i l_i/\sum_j c_j l_j$. Then the numbers of territories won by each type $\Delta_+ n_1,\Delta_+ n_2,\ldots,\Delta_+ n_G$ follow a multinomial distribution with $U$ trials and success probabilities $\frac{c_1 l_1}{\sum_j c_j l_j},\frac{c_2 l_2}{\sum_j c_j l_j},\ldots,\frac{c_G l_G}{\sum_j c_j l_j}$, respectively. Type $i$ is expected to win $c_i l_i/\sum_j c_j l_j$ of the $U$ available territories, and deviations from this expected outcome are small (since $T$ is large by assumption), giving 
\begin{equation}
\Delta_+ n_i=\frac{c_i l_i}{\sum_j c_j l_j}U=\frac{c_i l_i}{\overline{c}L}U, \label{eq:lottery}
\end{equation}
where $\overline{c}=\sum_j c_j m_j/M$ is the mean competitive ability for a randomly selected propagule, $L=M/U$ is the total propagule density and $M=\sum_j m_j$ is the total number of propagules. In section ``Analytical approximation of the density-dependent lottery'', we derive a generalization of Eq.~\eqref{eq:lottery} that accommodates arbitrary propagule densities $l_i$.

There is a close connection between the classic lottery model and the Wright-Fisher model of genetic drift \citep{svardal_2015}. In the Wright-Fisher model, type abundances are sampled each generation from a multinomial distribution with success probabilities $w_i n_i/\sum_j w_j n_j$, where $w$ is relative fitness and the $n_i$ are  type abundances in the preceding generation. Population size $N$ remains constant. This is equivalent to the classic lottery model with non-overlapping generations ($d_i=1$ for all $i$) and relative fitness given by $w_i=b_i c_i$ i.e. a product of fertility and viability \citep[pp. 185]{crow_1970}. Thus, the classic lottery model is essentially the Wright-Fisher model extended to allow overlapping generations, but ignoring drift. This means that our extension of the classic lottery model to arbitrary densities represents a variable-density generalization of the Wright-Fisher model (we also do not consider drift here).

@@

Eq.~\eqref{eq:master} takes a much simpler form if all types are competitively equivalent ($c_i=c$),
\begin{equation}
\Delta_+ n_i = \frac{l_i}{L}(1-e^{-L})U. \label{eq:masterequalc}
\end{equation}
Here $1-e^{-L}$ is the fraction of territories that receive at least one propagule under Poisson dispersal, $(1-e^{-L})U$ is the total number of territories gained, and type $i$ receives a fraction $l_i/L$ of these. Total population density thus grows according to
\begin{equation}
\Delta N=(1-e^{-L})U-\sum_i d_i n_i \label{eq:Nmaster}
\end{equation}

\subsection*{Selection-dependent density and $K$-selection}

Equipped with our variable-density lottery model, we now start evaluating the validity of Eq.~\eqref{eq:canonical}. In this section we explore whether we should expect population density to vary as a result of selection \citep{prout_1980}. Since the idea that density does vary with selection is closely connected to the notion of ``$K$-selection'', we start by revisiting MacArthur's analysis of selection in crowded environments \citep{macarthur_1967}.

MacArthur considers a population with two types that have densities $n_1$ and $n_2$ subject to density-dependent growth described by
\begin{equation}
\frac{d n_1}{d t}=f_1(n_1,n_2)\qquad\frac{d n_2}{d t}=f_2(n_1,n_2). \label{eq:macgeneral}
\end{equation}
The environment is assumed to remain constant apart from the type densities. The functions $f_1$ and $f_2$ must decline to zero if $n_1$ or $n_2$ are sufficiently large, because no population has unlimited resources. This defines the nullclines $f_1(n_1,n_2)=0$ and $f_2(n_1,n_2)=0$ in $(n_1,n_2)$ space. The outcome of selection is then determined by the relationship between these nullclines. Specifically, a type will be excluded if its nullcline is completely contained in the region bounded by the other type's nullcline. In other words, for a type to have the possibility of persisting, it must be able to keep growing to higher densities than the other type can tolerate in some region of $(n_1,n_2)$ space (Fig.~\ref{fig:Ksel}a).

@@

Thus, when MacArthur concludes that  ``fitness is $K$'' in crowded populations \citep[pp. 149]{macarthur_1967}, the meaning is that selection either favors the ability to keep growing at ever higher densities (moving a type's own nullcline outwards), or the ability to suppress the growth of competitors at lower densities (moving the nullcline of competitors inwards) \citep{gill_1974}. This general idea is much broader than ``$K$-selection'' in the sense of selection for greater saturation density, and applies even if the nullclines are nonlinear to such an extent that the ``$K$'' values themselves do not give much information about the regions of high-density growth.

@@

It is obvious from Eq.~\eqref{eq:LV} that selection can favor a superior competitor in a crowded population even if its saturation density is the same as, or lower than that of the other types present. However, note that the Lotka-Volterra model still couples selection to population density \citep{smouse_1976}. Fig.~\ref{fig:LVvslottery}a shows Lotka-Volterra selection between two types with the same saturation density ($\alpha_{11}=\alpha_{22}$, $\alpha_{21}>\alpha_{12}$). Even though the initial and final densities of a sweep are the same, density is not constant over a sweep. Only a highly restricted subset of $r$ and $\alpha$ values will keep $N$ constant over a selective sweep (further details in Appendix C). Intuitively, for one type to exclude another with the same saturation density, competitive suppression of growth between types must be stronger than competitive suppression of growth within types, causing a dip in $N$ over the sweep.

@@

Note that this behavior of $c$-selection arises directly from the role of $c$ as a trait determining relative competitive success in territorial contests, and as such this behavior is closely related to the classical appeal to reproductive excesses to justify Eq.~\eqref{eq:canonical}. By contrast, previous models of selection-independent density either used unusual models of competition \citep{kimura1969natural,nei1971fertility}, or made restrictive parameter choices in the Lotka-Volterra model (Appendix C; \citealt{smouse_1976,mallet_2012}).

@@

In the previous section we showed that selection and the regulation of population density can be completely independent of each other even if population growth is density-regulated, and moreover that MacArthur's general argument \citep{macarthur_1967} never precluded this possibility. Nevertheless, selection and density regulation \textit{are} intimately connected in widely used models of population growth. To understand why this poses a problem for Eq.~\eqref{eq:canonical}, consider the simple birth-death model \cite[pp. 20]{kostitzin_1939} 
\begin{equation}
\frac{d n_i}{dt}=(b_i -\delta_iN) n_i \label{eq:simplebirthdeath}
\end{equation}
where $\delta_i$ is the per-capita increase in type $i$'s mortality rate due to crowding (for simplicity, there are no deaths when uncrowded). Then, starting from a monomorphic population, the frequency of an adaptive $\delta$-variant $\delta_i\rightarrow \delta_i(1-\epsilon)$ obeys 
\begin{equation}
\frac{d p_i}{dt}=\epsilon \delta_i N p_i(1-p_i). \label{eq:Ndependentsweep}
\end{equation}
The selection coefficient $s=\epsilon \delta_i N$ thus depends on density (compare with Model III in \cite{kimura1969natural}). On the other hand, the frequency of an adaptive $b$-variant $b_i\rightarrow b_i(1+\epsilon)$ will exactly obey Eq.~\eqref{eq:canonical} with $s=\epsilon b_i$, independent of density.

@@

In particular, we emphasize that MacArthur's argument does not justify the widespread intuition that selection in crowded environments is necessarily connected to achieving greater densities. This intuition is largely an artifact of the models historically used in the density-dependent selection literature, which ignore the existence of relative contests.

@@

Another issue with the constant-$N$ relative fitness description of selection is that it precludes consideration of longer-term aspects of the interplay between evolution and ecology such as population extinction. A variety of approaches have been developed for dealing with these issues in quantitative genetics \citep{burger1995evolution,engen_2013}, population genetics \citep{bertram2017predicting} and adaptive dynamics \citep{ferriere2013eco,dieckmann2004adaptive}. Although this manuscript, and density-dependent selection more broadly, are pertinent to the interplay between ecology and evolution \citep{travis_2013}, our focus here is the description of the time-dependent process by which selection changes allele frequencies. This is particularly critical for making sense of evolution at the genetic level, for which we now have abundant data.
